fMRI by caveat

### 2023-03-19 paper.qmd
up to number 11

---
title: "fMRI by caveat"
author: "Tomas Knapen"
toc: true
number-sections: false
highlight-style: pygments
bibliography: lib/export_2023-3-16.bib
csl: templates/nature.csl
format:
  html:
    code-fold: true
  pdf:
    geometry:
        - top=30mm
        - left=0mm
  docx: default
jupyter: python3
---

@@

### Caveat I: MR Images are very different from photographs {#c1}

@@

### Caveat II: The scanner can image many things. BOLD is just one of those things. {#c2}

@@

### Caveat III: Things in an MR image may not be where they appear to be. {#c3}

@@

### Caveat IV: Blood, not neurons. {#c4}

@@

### Caveat V: Dissociations between BOLD and neuronal firing. {#c5}

@@

### Caveat VI: Yellow-red blobs are statistics, not 'activation'. {#c6}

@@

### Caveat VII: the BOLD response is not the only functional contrast mechanism that we can use to measure brain activations. {#c7}

@@

### Caveat VIII: Gradient-recalled echo imaging is very sensitive but less selective than other acquisitions. {#c8}

The "standard" BOLD imaging technique reverses spins after excitation using a change in the magnetic field gradient. We then capture RF signals after they re-align to form an 'echo'. Hence the name gradient-recalled echo (GRE) or gradient echo (GE). An older alternative technique, devised by Erwin Hahn in 1949, is called spin echo (SE). In this scheme, a 180$\degree$ RF pulse is used to reflect the magnetic spins, so that the return to form an RF 'echo'. The time at which the echo happens is double that at which the 180$\degree$ RF pulse was applied. SE is used a lot in anatomical imaging in the form of turbo-spin-echo sequences, in which continuous 180$\degree$ RF pulse reflections are used to speed up acquisition. For functional MRI, SE is less sensitive but more selective than GRE. We can use it to focus more on effect originating from smaller vessels, whereas GRE is often biased by signals originating from larger vessels. [mri-questions on GRE vs SE](https://mriquestions.com/gre-vs-se.html)

@@

### Caveat IX: The shape of the BOLD response differs a lot across voxels, individuals, and age. {#c9}

@@

### Caveat X: our use of the GLM depends on an LTI assumption. {#c10}

The use of the GLM depends on the assumption that the relation between the BOLD response and the thing of interest (neural responses) is a linear time-invariant (LTI) system. This means that doubling the strength of a neural signal will show a double amplitude response in the BOLD response. It also means that the response to a two-second event is equal to the sum of two one-second event responses, the second having been shifted by 1 second. Using these assumptions we can now use convolution with the impulse-response of the system (the HRF) to predict BOLD responses driven by specific events. Also, this means that we can gauge the strength of neural responses by looking at the $\beta$ weights that roll out of a GLM. It's important to realize that this is an _assumption_ that we cannot fully verify. It's clear that as a first pass it holds, though, but as opposed to simple electrical wirings, the complicated biological machinery driving the BOLD response is likely to show nonlinearities that break the LTI assumption. Just check out the examples above in Caveat @c5 to convince yourself that there are limits to the LTI assumption [@10.1038/nature07664].

### Caveat XI: GLM statistics depend on specific assumptions. {#c11}

The statistics we do on top of our GLMs assume independent, identically distributed samples - an assumption that is very common in more statistical applications. However, the noise that ends up as residuals actually has strong temporal autocorrelation. This means that the covariance structure of the noise doesn't follow an identity matrix, and that we should estimate this structure and use it to correct our statistics. If we do not take into account this correlational structure in our noise, we will overestimate the degrees of freedom that go into our T-statistic calculations, greatly inflating them and causing a lot of false positives. Similarly, cluster-based statistical procedures often use assumptions regarding the spatial autocorrelation of noise. If we assume such structure to be more narrow than it actually is, we run into problems [@10.1073/pnas.1602413113]. These problems, as in @c10, are that we will be too lenient and produce false-positive results.

>>>>

---
title: "fMRI by caveat"
author: "Tomas Knapen"
toc: true
number-sections: true
highlight-style: pygments
bibliography: lib/export_2023-3-16.bib
csl: templates/nature.csl
format:
  html:
    code-fold: true
  pdf:
    geometry:
        - top=30mm
        - left=30mm
        - top=30mm
        - left=30mm
  docx: default
jupyter: python3
---

@@

### Caveat I: MR Images are very different from photographs {#sec-c1}

@@

### Caveat II: The scanner can image many things. BOLD is just one of those things. {#sec-c2}

@@

### Caveat III: Things in an MR image may not be where they appear to be. {#sec-c3}

@@

### Caveat IV: Blood, not neurons. {#sec-c4}

@@

### Caveat V: Dissociations between BOLD and neuronal firing. {#sec-c5}

@@

### Caveat VI: Yellow-red blobs are statistics, not 'activation'. {#sec-c6}

@@

### Caveat VII: the BOLD response is not the only functional contrast mechanism that we can use to measure brain activations. {#sec-c7}

@@

### Caveat VIII: Gradient-recalled echo imaging is very sensitive but less selective than other acquisitions. {#sec-c8}

The "standard" BOLD imaging technique reverses spins after excitation using a change in the magnetic field gradient. We then capture RF signals after they re-align to form an 'echo'. Hence the name gradient-recalled echo (GRE) or gradient echo (GE). An older alternative technique, devised by Erwin Hahn in 1949, is called spin echo (SE). In this scheme, a 180$^{\circ}$ RF pulse is used to reflect the magnetic spins, so that the return to form an RF 'echo'. The time at which the echo happens is double that at which the 180$^{\circ}$ RF pulse was applied. SE is used a lot in anatomical imaging in the form of turbo-spin-echo sequences, in which continuous 180$^{\circ}$ RF pulse reflections are used to speed up acquisition. For functional MRI, SE is less sensitive but more selective than GRE. We can use it to focus more on effect originating from smaller vessels, whereas GRE is often biased by signals originating from larger vessels. [mri-questions on GRE vs SE](https://mriquestions.com/gre-vs-se.html)

@@

### Caveat IX: The shape of the BOLD response differs a lot across voxels, individuals, and age. {#sec-c9}

@@

### Caveat X: our use of the GLM depends on an LTI assumption. {#sec-c10}

The use of the GLM depends on the assumption that the relation between the BOLD response and the thing of interest (neural responses) is a linear time-invariant (LTI) system. This means that doubling the strength of a neural signal will show a double amplitude response in the BOLD response. It also means that the response to a two-second event is equal to the sum of two one-second event responses, the second having been shifted by 1 second. Using these assumptions we can now use convolution with the impulse-response of the system (the HRF) to predict BOLD responses driven by specific events. Also, this means that we can gauge the strength of neural responses by looking at the $\beta$ weights that roll out of a GLM. It's important to realize that this is an _assumption_ that we cannot fully verify. It's clear that as a first pass it holds, though, but as opposed to simple electrical wirings, the complicated biological machinery driving the BOLD response is likely to show nonlinearities that break the LTI assumption. Just check out the examples above in Caveat @sec-c5 to convince yourself that there are limits to the LTI assumption [@10.1038/nature07664].

### Caveat XI: GLM statistics depend on specific assumptions. {#sec-c11}

The statistics we do on top of our GLMs assume independent, identically distributed samples - an assumption that is very common in more statistical applications. However, the noise that ends up as residuals actually has strong temporal autocorrelation. This means that the covariance structure of the noise doesn't follow an identity matrix, and that we should estimate this structure and use it to correct our statistics. If we do not take into account this correlational structure in our noise, we will overestimate the degrees of freedom that go into our T-statistic calculations, greatly inflating them and causing a lot of false positives. Similarly, cluster-based statistical procedures often use assumptions regarding the spatial autocorrelation of noise. If we assume such structure to be more narrow than it actually is, we run into problems [@10.1073/pnas.1602413113]. These problems, as in @sec-c10, are that we will be too lenient and produce false-positive results.
